---
title: "Testing Functions"
teaching: 30
exercises: 0
questions:
- "Why should we test our functions?"
- "How do we test our functions?"
objectives:
- "To use testthat to run unit tests on our functions"
keypoints:
- "Use testthat to formalise your tests"
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-testing-R-")

```

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
load("checkpoints/02end.RData")
```

At the end of the previous episode we verified that our `cleanfields()` function worked on one of the variables by plotting a graph of the cleaned data.  This showed that all the `-999.9` values had been recoded to `NA`.

We can (and should) adopt a more formal approach to testing the functionality of our function by writing *unit tests*.   These let us test *specific* elements of our function's functionality.    If we need to modify our function the unit tests let us confirm that our change hasn't broken any of the function's existing functionality.


For example, let's test that our `cleanfield()` function behaves as we expect:

```{r}
testvector <- c(1,2,-999.99)
cleanfield(testvector)

```

The power of testing comes when we write code to *check* that the function has behaved as expected.  For example:

FIXME - this is a horrid example, as it takes us down the rabbit-hole of dealing with `NA`s in logic tests
and the all.equal() wrinkle

```{r}
expectedResults <- c(1,2,NA)

if( isTRUE(all.equal(cleanfield(testvector), expectedResults)) ){
  print("Test passed")
} else {
  print("Test failed")
}

```


Writing such tests is cumbersome.  The R package `testthat` makes it easy for us to write and run tests to verify our function behaves as we'd like.  If you put your functions in a package (see, for example, FIXME), it integrates well with the development process.  For now let's put our tests in a separate file.  In this file we'll load the `testthat` package.  An example of a file containing tests is included in the data you downloaded at the start of the course.

```{r, eval=FALSE}
context("Cleaning fields")

test_that("Can clean a field", {
  
  testvector <- c(1,2,-999.99)
  
  expect_equal(c(1,2,NA), cleanfield(testvector))
  expect_equal(c(1,NA,-999.99), cleanfield(testvector, missingvalue = 2))
  
  expect_equal(NA, cleanfield(-999.99))
  expect_equal(1, cleanfield(1))
})

test_that("Can clean multiple fields",{
  
  testtibble <- tibble(a = c(1,2,3), b = c(3,4,-999.99), c = c(1,-999.99,2))
  
  cleanedtibbleSinglefield <- tibble(a = c(1,2,3), b = c(3, 4, NA), c = c(1, -999.99, 2))
  expect_equal(cleanedtibbleSinglefield, cleanfields(testtibble, "b"))
  
  cleanedtibbleMultifield <- tibble(a = c(1,2,3), b = c(3, 4, NA), c = c(1, NA, 2))
  expect_equal(cleanedtibbleMultifield, cleanfields(testtibble, c("b","c")))
})

```

We can execute the tests by loading the `testthat` package, and running `test_file()`:

```{r}
library("testthat")
test_file("tests/test_cleandata.R", env=.GlobalEnv)
```

To show you what happens when a test fails:

```{r}
test_file("tests/test_fail.R")
```


## Challenge

The `test_cleandata.R` file doesn't test whether the code we wrote to check variable names exist
works properly.  Modify the file where indicated to test this functionality.  You may find the `expect_error()` 
function useful.

## Solution

```{r}

test_that("Missing variable detection works",{
  
  # Make a test tibble to run the tests on
  testtibble <- tibble(a = c(1,2,3), b = c(4,5,6) )
  # Test that we get an error if we specify a missing variable name
  expect_error(cleanfields(testtibble, c("a","d")))
  
})
```


## Challenge: Repeated variable names

At present we can pass the same variable name to our `cleanfields()` function more than once, e.g.
```{r, eval=FALSE}
cleanfields(mydata, c("field1","field2","field1"))
```

The function will still work, but the second pass through the data for the repeated field will have no effect.
The user probably didn't mean to repeat the field; perhaps they made a typing error and meant field3. 

Write a test to expect a warning if a variable name is repeated.  Run your tests, and check that the test failure is reported.

Then modify your function so that the test passes.   Developing code in this way is referred to as test driven development. FIXME - provide a link to good resource.  Ask Rob?

Hint:  There are a number of ways of checking whether a vector contains duplicated elements.  `unique()` returns only the unique elements of a vector.  You could compare the length of this vector to the original vector.  `anyDuplicated()` returns the index of the first duplicated element in a vector, or 0 if there are no duplicate elements.

## Solution

A suitable test would be:
```{r, eval=FALSE}

test_that("Duplicate variables give a warning",{
  
  testtibble <- tibble(a = c(1,2,3), b = c(4,5,6) )
  
  expect_warning(cleanfields(testtibble, c("a","a")))
  # Do we still get a warning if the duplicates aren't adjacent?
  expect_warning(cleanfields(testtibble, c("a","b","a")))
})
```

These tests fail unless we modify our `cleanfields()` function:

```{r}
cleanfields <- function(dataset, fieldlist){
  
  if ( !all(fieldlist %in% names(dataset)) ) {
    stop("Attempting to clean variables that do not exist in the dataset")
  }
  
  if ( anyDuplicated(fieldlist) != 0 ) {
    warning("Duplicated variable names specified")
  }
  
  for (f in fieldlist) {
    dataset[[f]] <- cleanfield(dataset[[f]])
  }
  
  return(dataset) 
}

```




