---
title: "Loading multiple files"
teaching: 30
exercises: 0
questions:
- "Example ?"
objectives:
- "Example objective"
keypoints:
- "Example keypoint"
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-multi-R-")

```

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
load("checkpoints/04end.RData")
```

## Loading more than one file at once

We have many years of weather data, and we would like to load them into a single tibble.  We can modify our `loadWeatherData()` function to do this.  

We already know how to iterate over a vector using a `for` loop.  Let's make an example vector by hand, while we develop the function:

```{r}
weatherfiles <- c("data/met_mlo_insitu_1_obop_hour_1977.txt", "data/met_mlo_insitu_1_obop_hour_1978.txt", "data/met_mlo_insitu_1_obop_hour_1979.txt")
```

This is what we want our function to do::

```{r}
loadWeatherDataPseudoCode <- function(weatherfiles){
  for (f in weatherfiles) {
    print(paste("Load in data for file", f))
    print("Append the data to the data we've already loaded")
  }
  print("return the tibble containing all the data")
}

loadWeatherDataPseudoCode(weatherfiles)

```

## Create a new function or modify an existing one?

We could decide to create a new function which would call `loadWeatherData()` for each file,
rather than modifying our existing function.  There are pros and cons to extending the existing
function, or creating a new one.  MORE HERE - ease of debugging vs single file is just file vector 
with length 1

{: .callout}


```{r}
loadWeatherData <- function(infiles){
  # Load in a weather data file
  
  for (infile in infiles) {  
    weather <- read_table(infile,
                          col_names = c("obs",
                                        "yyyy",
                                        "mm",
                                        "dd",
                                        "hh",
                                        "winddir",
                                        "windspeed",
                                        "windsteadiness",
                                        "pressure",
                                        "temperature2m",
                                        "temperature10m",
                                        "temperaturetop",
                                        "relhumidity",
                                        "precipitation" ),
                          col_types = cols(
                            obs = col_character(),
                            yyyy = col_integer(),
                            mm = col_character(),
                            dd = col_character(),
                            hh = col_character(),
                            winddir = col_integer(),
                            windspeed = col_double(),
                            windsteadiness = col_integer(),
                            pressure = col_double(),
                            temperature2m = col_double(),
                            temperature10m = col_double(),
                            temperaturetop = col_double(),
                            relhumidity = col_integer(),
                            precipitation = col_integer()
                          )
    )
    
    
    weather <- weather %>% mutate(recdate = lubridate::ymd_h(paste(yyyy,mm,dd,hh)))
    
    missingvalues <- c(winddir = -999,
                       windspeed = -999.9,
                       windsteadiness = -9,
                       pressure = -999.9,
                       temperature2m = -999.9,
                       temperature10m = -999.9,
                       temperaturetop = -999.9,
                       relhumidity = -99,
                       precipitation = -99)
    weather <- cleanfields(weather, missingvalues)
  }
  return(weather)
}
```


```{r}
cleanweather <- loadWeatherData(weatherfiles)
```

That looks like it's worked; but what does our `cleanweather` data set contain?

```{r}
cleanweather %>%  
  group_by(yyyy) %>% 
  count()  
```

We only have data for the most recent year that we read in.  This is because the `weather` data-set gets overwritten each time we run through the `for` loop.  Let's modify the function so that we append the current year's data to a tibble that we define out of the loop.

**This is a really bad thing to do in R - should make a list and bind togeher**


```{r}
loadWeatherData <- function(infiles){
  # Load in a weather data file
  allweather <- NULL
  for (infile in infiles) {  
    weather <- read_table2(infile,
                          col_names = c("obs",
                                        "yyyy",
                                        "mm",
                                        "dd",
                                        "hh",
                                        "winddir",
                                        "windspeed",
                                        "windsteadiness",
                                        "pressure",
                                        "temperature2m",
                                        "temperature10m",
                                        "temperaturetop",
                                        "relhumidity",
                                        "precipitation" ),
                          col_types = cols(
                            obs = col_character(),
                            yyyy = col_integer(),
                            mm = col_character(),
                            dd = col_character(),
                            hh = col_character(),
                            winddir = col_integer(),
                            windspeed = col_double(),
                            windsteadiness = col_integer(),
                            pressure = col_double(),
                            temperature2m = col_double(),
                            temperature10m = col_double(),
                            temperaturetop = col_double(),
                            relhumidity = col_integer(),
                            precipitation = col_integer()
                          )
    )
    
    
    weather <- weather %>% mutate(recdate = lubridate::ymd_h(paste(yyyy,mm,dd,hh)))
    
    missingvalues <- c(winddir = -999,
                       windspeed = -99.9, ## Some records use this as missing
                       windspeed = -999.9, ## Others use this(!)
                       windsteadiness = -9,
                       pressure = -999.9,
                       temperature2m = -999.9,
                       temperature10m = -999.9,
                       temperaturetop = -999.9,
                       relhumidity = -99,
                       precipitation = -99)
    weather <- cleanfields(weather, missingvalues)
    
    
    
    
    allweather <- bind_rows(allweather, weather)
  }
  return(allweather)
}
```




```{r}
cleanweather <- loadWeatherData(weatherfiles)
```


```{r}
cleanweather %>%  
  group_by(yyyy) %>% 
  count()  
```

So we can now pass more than one file to our `loadWeatherData` function.   Rather than type all the files in, we
can use the `list.files()` function to generate the vector of filenames:

```{r}
weatherfiles <- list.files(path="./data", "met_mlo_ins*",full.names=TRUE)
cleanweather <- loadWeatherData(weatherfiles)
```

```{r}
summary(cleanweather)
```


```{r}
cleanweather %>% 
  ggplot(aes(x=recdate, y=temperature2m)) + geom_line()
```

```{r}
cleanweather %>% 
  filter(hh == "00") %>% 
  group_by(yyyy,mm) %>% 
  summarise(meantemp = mean(temperature2m, na.rm=TRUE)) %>% 
  ggplot(aes(x=mm,y=yyyy,fill=meantemp)) + geom_raster()
```

