---
title: Functions and dplyr
teaching: 45
exercises: 15
questions:
- "How do we write functions that work with dplyr?"
objectives:
- "Understand the benefits and drawbacks of non standard evaluation"
keypoints:
- "FIXME - keypoints go here"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
library(tidyverse)
knitr_fig_path("06-")
# Silently load in the data so the rest of the lesson works
gapminder <- read_csv("data/gapminder-FiveYearData.csv")
```

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-multi-R-")

```
```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
load("checkpoints/05end.RData")
```


A lot of the "cleverness" of the tidyverse comes from how it handles the values we pass to its functions.
This lets us write code in a very expressive way (for example, the idea of treating the dplyr
pipelines as a sentence consisting of a series of verbs (such as `filter()` and `mutate()`), with the `%>%` operator being read as "and then".)

Unfortunately this comes at a price when we come to use `dplyr` in our own functions.  The `dplyr` functions use what is known as "Non Standard Evaluation"; this means that they don't evaluate all of their parameters in the standard way that the examples in this episode have used so far.  This is probably the most complicted concept we will cover today, so pleae don't worry if you don't "get" it at first attempt. 

We'll illustrate this by way of an example.

Let's write a function that will filter our weather data to only keep observations when it is both warm (which we'll define as at least 18 degrees C), and windy (which we'll define as a wind speed of at least 6m/s).
First, let's think about how we'd do this *without* using a function.  We'll then put our 
code in a function; this will let us easily repeat the calculation for other data-sets, without re-writing the code.

```{r}
warmWindy <- cleanweather %>% 
  filter(windspeed >= 6, temperature2m >= 18)

```

## Challenge:

In how many observations is it both warm and windy?

## Solution:

```{r}
nrow(warmWindy)
```

It is only warm and windy `r nrow(warmWindy)` times in the entire data set.

{: .solution}
{: .challenge}

Let's put this code in a function; note that we change the input data to be the parameter we pass when we call
the function, so that we can adjust the wind and temperature thresholds (it's _really_ easy to forget to make these changes when you start making functions from existing code):

```{r}
getWarmWindyObservations <- function(indata, windspeed = 6, temperature2m = 18){
  
  warmWindy <- indata %>% 
    filter(windspeed >= windspeed, temperature2m >= temperature2m)
  
  return(warmWindy)
  
}
```

But if we run this function, it doesn't do what we'd expect:

```{r}
getWarmWindyObservations(cleanweather)
```

All of our data is returned.

Look at the line: 

```{r, eval=FALSE}
    filter(windspeed >= windspeed, temperature2m >= temperature2m) %>% 
```

We passed the values of `windspeed` and `temperature2m` into the function when we called it.  R has no way of knowing we're referring to the function parameter's `windspeed` and `temperature2m` variables rather than the tibble's `windspeed` and `temperature2m` variables.     We run into this problem because of dplyr's "Non Standard Evaluation" (NSE); this means that the functions don't follow R's usual rules about how parameters are evaluated.  

NSE is usually really useful; when we've written things like `cleanweather %>% select(recdate, tempeature2m)` we've made use of non standard evaluation.  This is much more intuitive than the base R equivalent, where we'd have to write something like `cleanweather[, names(cleanweather) %in% c("recdate", "temperature2m")]`. Non standard evaluation lets the `select()` function work out that we're referring to variable names in the tibble.  Unfortunately this simplicity comes at a price when we come to write functions.  It means we need a way of telling R whether we're referring to a variable in the tibble, or a parameter we've passed via the function.

If we're using standard evaluation (with "regular" R functions), then we see the value of the parameters we've passed in.  For example, let's put
a couple of `print()` statements in our function:


```{r}

getWarmWindyObservations <- function(indata, windspeed = 6, temperature2m = 18){
  #For debugging
  print(windspeed)
  print(temperature2m)
  
  warmWindy <- indata %>% 
    filter(windspeed >= windspeed, temperature2m >= temperature2m)
  
  return(warmWindy)
  
}

warmWindy <- getWarmWindyObservations(cleanweather)

```

As you'd expect, we get the (default) parameter values that we defined with the function.
We can use our parameters like normal variables in our function _except_ when we're using it as part of a parameter to a `dplyr` verb (e.g. `filter`), or another function that uses non standard evaluation.   We need to _unquote_ the parameters so that the `dplyr` function can see their contents (i.e. 6 and 18 in this example).  We do this using the `UQ()` ("unquote") function:

```{r, eval=FALSE}
    filter(windspeed >= UQ(windspeed), temperature2m >= UQ(temperature2m) ) %>% 
```

When the filter function is evaluated it will see:


```{r, eval=FALSE}
    filter(windspeed >= 6, temperature2m >= 18 ) %>% 
```

Modifying our function, we have:

```{r}
getWarmWindyObservations <- function(indata, windspeed = 6, temperature2m = 18){
  
  warmWindy <- indata %>% 
    filter(windspeed >= UQ(windspeed), temperature2m >= UQ(temperature2m))
  
  return(warmWindy)
  
}
getWarmWindyObservations(cleanweather)
                 

```

Our function now works as expected.

> ## Another way of thinking about quoting
> 
> (This section is based on [programming with dplyr](http://dplyr.tidyverse.org/articles/programming.html))
>
> Consider the following code:
>
> ```{r}
> greet <- function(person){
>   print("Hello person")
> }
> 
> greet("David")
> ```
> 
> The `print` function has no way of knowing that `person` refers to the variable `person`, and isn't 
> part of the string `person`.  To make the contents of the `person` variable visible to the function we
> need to construct the string, using the `paste` function:
>
> ```{r}
> greet <- function(person){
>   print(paste("Hello", person))
> }
> 
> greet("David")
> ```
> This means that the `person` variable is evaluated in an _unquoted_ environment, so its contents can be evaluated.
{: .callout}

There is one small issue that remains; how does filter _know_ that the first `windspeed` in ``` filter(windspeed >= UQ(windspeed), ...``` refers to the `windspeed` variable in the tibble? (and similarly for the `temperature2m` variable). What happens if we delete the  year variable? Surely this should give an error?

```{r}
weather_nowind <- cleanweather %>% select(-windspeed)
getWarmWindyObservations(weather_nowind)
```

As you can see, it doesn't; instead the `filter()` function will "fall through" to look for the `windspeed` variable in `filter()`'s _calling environment_,  This is the `getWarmWindyObservations()` environment, which does have a `windspeed` variable.  Since this is a standard R variable, it will be implicitly unquoted, so `filter()` will see:

```{r, eval=FALSE}
    filter(windspeed >= windspeed) %>% 
```

which is always `TRUE`, so the filter won't do anything!

We need a way of telling the function that the first `windspeed` "belongs" to the data.  We can do this with the `.data` pronoun:

```{r}
getWarmWindyObservations <- function(indata, windspeed = 6, temperature2m = 18){
  
  warmWindy <- indata %>% 
    filter(.data$windspeed >= UQ(windspeed), .data$temperature2m >= UQ(temperature2m))
  
  return(warmWindy)
  
}

# This works
getWarmWindyObservations(cleanweather)

# When we're missing a variable we get an error
getWarmWindyObservations(weather_nowind)

```                 

As you can see, we've also used the `.data` pronoun when getting the warm and windy observations. This tells R that the `windspeed` and `temperature2m` variables are in the tibble that we're working on, and prevents R looking into the calling environment. If the variables aren't found in the data, we get an error. 

## Challenge: Calculating the average wind speed per day

Create a new function that will take a tibble containing weather data, and two parameters,
`fromyear` and `toyear`.   The function should filter the data to include only data between the two years
specified, and calculate the average wind speed per day for these data points.

Hint:  The `between()` function in dplyr may be useful.

## Solution

```{r}
averageWindSpeed <- function(indata, fromyear, toyear){
  
  avgwind <- indata %>% 
    filter(between(.data$yyyy, UQ(fromyear), UQ(toyear))) %>% 
    group_by(.data$yyyy, .data$mm, .data$dd) %>% 
    summarise(avgwindspeed = mean(windspeed))
  
  return(avgwind)
}

averageWindSpeed(cleanweather, 1980, 1981)
```

Note that we use `UQ()` when we evaluate `fromyear` and `toyear`, and use the `.data` pronoun to make it clear which variables 
should be in our data.

{: .solution}
{: .challenge}


> ## Tip
>
> The "programming with dplyr" vignette is highly recommended if you are
> writing functions for dplyr.  This can be accessed by typing `vignette("programming", package="dplyr")`
>  
{: .callout}

[man]: http://cran.r-project.org/doc/manuals/r-release/R-lang.html#Environment-objects
[chapter]: http://adv-r.had.co.nz/Environments.html
[adv-r]: http://adv-r.had.co.nz/


In this episode we've introduced the idea of writing your own functions, and talked about the
complications caused by non standard evaluation.   

[roxygen2]: http://cran.r-project.org/web/packages/roxygen2/vignettes/rd.html
[testthat]: http://r-pkgs.had.co.nz/tests.html
